---
title: "Cyclistic Analysis"
author: "Dakota Stewart"
date: "2022-08-19"
output: 
  html_document: 
    theme: readable
data source: "https://divvy-tripdata.s3.amazonaws.com/index.html"
---

##### Package installation and library activation.

```{r}
install.packages("tidyverse")

install.packages("dplyr")

install.packages("janitor")

install.packages("lubradate")

install.packages("ggplot2")

library(tidyverse)

library(dplyr)

library(janitor)

library(lubridate)

library(ggplot2)
```

##### Setting the working directory.

`setwd("~/Data analytics/capstone/Total_2021_Divvy_Tripdata")`

##### Installation of my data.

```{r}
Jan_Tripdata<- read.csv("202101-divvy-tripdata.csv") 

Feb_Tripdata<- read.csv("202102-divvy-tripdata.csv") 

Mar_Tripdata<- read.csv("202103-divvy-tripdata.csv") 

Apr_Tripdata<- read.csv("202104-divvy-tripdata.csv") 

May_Tripdata<- read.csv("202105-divvy-tripdata.csv") 

Jun_Tripdata<- read.csv("202106-divvy-tripdata.csv") 

Jul_Tripdata<- read.csv("202107-divvy-tripdata.csv") 

Aug_Tripdata<- read.csv("202108-divvy-tripdata.csv") 

Sep_Tripdata<- read.csv("202109-divvy-tripdata.csv") 

Oct_Tripdata<- read.csv("202110-divvy-tripdata.csv") 

Nov_Tripdata<- read.csv("202111-divvy-tripdata.csv") 

Dec_Tripdata<- read.csv("202112-divvy-tripdata.csv")
```

##### Merging the 12 data frames into 1 data frame.

```{r}
Total_Tripdata_2021<- rbind(Jan_Tripdata,Feb_Tripdata,Mar_Tripdata,Apr_Tripdata,May_Tripdata,Jun_Tripdata,Jul_Tripdata,Aug_Tripdata,Sep_Tripdata,Oct_Tripdata,Nov_Tripdata,Dec_Tripdata)
```

##### Cleaning empty data with janitor.

```{r}
Total_Tripdata_2021<- janitor::remove_empty(Total_Tripdata_2021,which = c("rows","cols"))
```

##### Split the start date and time column into individual columns.

```{r}
Total_Tripdata_2021$date <- as.Date(Total_Tripdata_2021$started_at) Total_Tripdata_2021$month <- format(as.Date(Total_Tripdata_2021$date), "%m") Total_Tripdata_2021$day <- format(as.Date(Total_Tripdata_2021$date), "%d") Total_Tripdata_2021$year <- format(as.Date(Total_Tripdata_2021$date), "%Y") Total_Tripdata_2021$day_of_week <- format(as.Date(Total_Tripdata_2021$date), "%A")
```

##### Calculate Ride duration per user

```{r}
Total_Tripdata_2021$Ride_Duration<- difftime(Total_Tripdata_2021$ended_at,Total_Tripdata_2021$started_at, units = "secs")
```

##### Change the format from factor to nemeric

```{r}
is.factor(Total_Tripdata_2021$Ride_Duration) Total_Tripdata_2021$Ride_Duration <- as.numeric(as.character(Total_Tripdata_2021$Ride_Duration )) is.numeric(Total_Tripdata_2021$Ride_Duration )
```

##### Clean the dataset more to remove any additional null values and create a new dataset

```{r}
Total_Tripdata_2021_v2 <- Total_Tripdata_2021[!(Total_Tripdata_2021$start_station_name == "HQ QR" | Total_Tripdata_2021$Ride_Duration>5),]
```

##### Run a summary command to collect the mean,median,max, and min data

```{r}
mean(Total_Tripdata_2021_v2$Ride_Duration) 
median(Total_Tripdata_2021_v2$Ride_Duration) 
max(Total_Tripdata_2021_v2$Ride_Duration) 
min(Total_Tripdata_2021_v2$Ride_Duration) 
summary(Total_Tripdata_2021_v2$Ride_Duration)
```

##### Compare both members and casual users

```{r}
aggregate(Total_Tripdata_2021_v2$Ride_Duration ~ Total_Tripdata_2021_v2$member_casual, FUN = mean) 
aggregate(Total_Tripdata_2021_v2$Ride_Duration ~ Total_Tripdata_2021_v2$member_casual, FUN = median) 
aggregate(Total_Tripdata_2021_v2$Ride_Duration ~ Total_Tripdata_2021_v2$member_casual, FUN = max) 
aggregate(Total_Tripdata_2021_v2$Ride_Duration ~ Total_Tripdata_2021_v2$member_casual, FUN = min)
```

##### Order the days of the week so that they are easier to read

```{r}
Total_Tripdata_2021_v2$day_of_week <- ordered(Total_Tripdata_2021_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

##### Calculate the average ride duration by members and casual users

```{r}
aggregate(Total_Tripdata_2021_v2$Ride_Duration ~ Total_Tripdata_2021_v2$member_casual + Total_Tripdata_2021_v2$day_of_week, FUN = mean)
```

##### Disable scientific notation for our visualization

```{r}
options(scipen = 999)
```

##### Create a visualization to display total rides and the days of the week

```{r}
Total_Tripdata_2021_v2 %>% 

mutate(weekday = wday(started_at, label = TRUE)) %>% 

group_by(member_casual, weekday) %>% 

summarise(number_of_rides = n() ,average_duration = mean(Ride_Duration)) %>% arrange(member_casual, weekday)  %>% 

ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) + geom_col(position = "dodge")+  xlab("Days of the Week")+ylab("Number of Rides ") + scale_fill_discrete(name= "Cyclistic Users") + labs(title = "Total Number of Rides in 2021",subtitle = "By Days of the Week")
```

##### Create a graph to display average ride duration for members and casual users

```{r}
Total_Tripdata_2021_v2 %>%

 mutate(weekday = wday(started_at, label = TRUE)) %>% 

group_by(member_casual, weekday) %>%

 summarise(number_of_rides = n() ,average_duration = mean(Ride_Duration)) %>% arrange(member_casual, weekday)  %>%

 ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) + geom_col(position = "dodge") +  xlab("Days of the Week")+ylab("Ride Duration") + scale_fill_discrete(name= "Cyclistic Users") + labs(title = "Average Ride Duration",subtitle = "In Seconds")
```

##### Create a exportable table to upload to tablue

```{r}
Average_Tripdata_2021 <- aggregate(Total_Tripdata_2021_v2$Ride_Duration ~ Total_Tripdata_2021_v2$member_casual + Total_Tripdata_2021_v2$day_of_week, FUN = mean)
```

##### View the exportable table

```{r}
View(Average_Tripdata_2021)
```
